---
title: "Digging for Gold"
author: "Perry Stephenson"
date: "2018-05-01"
output:
  html_document: 
    toc: true
  html_notebook: default
---

```{r load_packages, results="hide", message=FALSE}
# devtools::install_github("perrystephenson/goldminer")
library(tidyverse)
library(goldminer) # https://github.com/perrystephenson/goldminer
```

```{r config, results="hide", message=FALSE}
config <- list()
params <- list()
env1 <- list()
env2 <- list()
config$rounds <- 1e6L # 1 million rounds
config$years <- 5
set.seed(12345)
```

# Approach

This analysis steps through several elements of the model, starting from the 
environment (which the operator of the mine cannot influence), through to some 
basic decision making strategies, robust modelled decisions, then some deep 
dives into specific decisions as guided by the data.

Accordingly, each of the environment variables will be modelled based on the
provided parameters for the full five year period regardless of whether or not
they are required in any individual simulation. Once the environment has been
modelled, the impacts of decisions made by the business can be reasoned about
without altering the modelled environment.

# Environment

### Gold Price

As specified, the initial price of gold in the first year is $1700, and the 
annual (absolute) change in the price is between -100 and 500, with an expected
value of 100.

```{r gold_price_config, results="hide", message=FALSE}
params$gold_price_init <- 1700
params$gold_price_delta <- list(a = -100, b = 500, c = 100)
```

```{r gold_price}
env1$gold_price <- gen_gold_price(rounds = config$rounds, 
                                  years = config$years, 
                                  gold_price_init = params$gold_price_init,
                                  gold_price_delta = params$gold_price_delta)
```

```{r plot_gold_price, echo=FALSE, fig.height=2}
env1$gold_price %>%
  gather("Year", "Price") %>%
  filter(Year != "Year1") %>% 
  ggplot() +
  geom_density(aes(x = Price, fill = Year, colour = Year), alpha = 0.6) +
  geom_line(data = data.frame(x=c(1700,1700),y=c(-Inf,Inf)), aes(x,y), col = "red") +
  labs(title = "Gold Price Distribution", 
       x = "Price (dollars)",
       y = "Probability Density")
```

The distribution for Year 1 is marked by a single line at $1700, given that it 
is fixed by definition.

### Gold Production

As specified, the initial gold production in the first year is between 300 and
1400 ounces, with an expected value of 500 ounces. The production in each 
subsequent year is determined using the previous year's output and multiplying
the output by a value with a range of 0.8 to 1.8 with an expected value of 1.4.

```{r gold_prod_config, results="hide", message=FALSE}
params$prod_init <- list(a = 300, b = 1400, c = 500)
params$prod_delta <- list(a = 0.8, b = 1.8, c = 1.4)
```

```{r gold_prod}
env1$gold_prod <- gen_gold_prod(rounds = config$rounds, 
                                years = config$years, 
                                prod_init = params$prod_init,
                                prod_delta = params$prod_delta)
```

```{r plot_gold_prod, echo=FALSE, fig.height=2}
env1$gold_prod %>%
  gather("Year", "Production") %>%
ggplot(aes(x = Production, fill = Year, colour = Year)) +
  geom_density(alpha = 0.6) +
  labs(title = "Gold Production Distribution", 
       x = "Production (ounces)",
       y = "Probability Density")
```

The production of gold is trending up over time, with the length of the positive
tail increasing each year. The uncertainty also increases over time.

### Leased Mine Income Share

The company owns a second mine that they lease to a third party. The contract is
structured in such a way that the income received includes a profit share
component. The range of income from leasing this second mine is between 1
million and 3 million dollars, with an expected value of 2 million dollars.

Each subsequent year the income from the lease is determined by modifying the
amount of the lease that was received in the **first year**. The income from the
mine is in the range of -20% (i.e. 0.8x multiplier) and 45% (i.e. 1.45x
multiplier) with an expected value of 10% (i.e. 1.1x multiplier).


```{r lease_profit_config, results="hide", message=FALSE}
params$lease_profit_init <- list(a = 1000000, b = 3000000, c = 2000000)
params$lease_profit_delta <- list(a = 0.8, b = 1.45, c = 1.1)
```

```{r lease_profit}
env1$lease_profit <- gen_lease_profit(rounds = config$rounds, 
                                      years = config$years, 
                                      profit_init = params$lease_profit_init,
                                      profit_delta = params$lease_profit_delta)
```

```{r plot_lease_profit, echo=FALSE, fig.height=2}
env1$lease_profit %>%
  gather("Year", "Profit") %>%
ggplot(aes(x = Profit, fill = Year, colour = Year)) +
  geom_density(alpha = 0.6) +
  labs(title = "Lease Profit Distribution", 
       x = "Lease Profit (dollars)",
       y = "Probability Density")
```

The probability distribution for each of the subsequent years is equal for all
intents and purposes due to the fact that years 2 through 5 are based on year 1
rather than the previous year.

### Reputation Management

The company incurs reputation management costs while they own the second 
(leased) mine. These costs have a fixed component ($350,000 per year whilst the
leased mine is still in operation) and a variable component ($2,500 per day) 
with an estimated range of 75 to 350 days per year (expected value 125 days).

```{r rep_mgmt_config, results="hide", message=FALSE}
params$rep_mgmt_fixed <- 350000
params$rep_mgmt_rate <- 2500
params$rep_mgmt_var <- list(a = 75, b = 350, c = 125)
```

```{r rep_mgmt}
env1$rep_mgmt <- gen_rep_mgmt(rounds = config$rounds, 
                              years = config$years, 
                              fixed = params$rep_mgmt_fixed,
                              rate = params$rep_mgmt_rate,
                              hours = params$rep_mgmt_var)
```

```{r plot_rep_mgmt, echo=FALSE, fig.height=2}
env1$rep_mgmt %>%
  gather("Year", "Costs") %>%
ggplot(aes(x = Costs, fill = Year, colour = Year)) +
  geom_density(alpha = 0.6) +
  labs(title = "Reputation Management Cost Distribution", 
       x = "Costs",
       y = "Probability Density")
```

The probability distribution for each year of reputation management is equal for
all intents and purposes due to the fact that the assumptions are independent
from year to year.

### Mine Sale

The leased mine can be sold at the start of any year, and the price offered will
be fixed as follows:

* Year 1 - $3,000,000
* Year 2 - $1,700,000
* Year 3 - $1,250,000
* Year 4 - $750,000
* Year 5 - $500,000

```{r sale_price, echo=FALSE}
params$sale_price <- list(
  "Year1" = 3000000,
  "Year2" = 1700000,
  "Year3" = 1250000,
  "Year4" =  750000,
  "Year5" =  500000) 
```

### Interest Rate

The interest rate is variable, but fixed for each 12 month period. The interest
rate is not known in advance, but is understood to be in the range of 8% to 16%
with an expected value of 12.75%.

```{r interest_rate_config, results="hide", message=FALSE}
params$interest_rate_range <- list(a = 0.08, b = 0.16, c = 0.1275)
```

```{r interest_rate}
env1$interest_rate <- gen_independent(rounds = config$rounds, 
                                        years = config$years, 
                                        t = params$interest_rate_range)
```

```{r plot_interest_rate, echo=FALSE, fig.height=2}
env1$interest_rate %>%
  gather("Year", "Rate") %>%
ggplot(aes(x = Rate, fill = Year, colour = Year)) +
  geom_density(alpha = 0.6) +
  labs(title = "Interest Rate Distribution", 
       x = "Interest Rate",
       y = "Probability Density")
```

The probability distribution for the interest rate for each year is equal for
all intents and purposes due to the fact that the assumptions are independent
from year to year.

### Selling Cost

There is a cost associated with selling the gold such as transportation, broker
fees, taxes and other charges. This is simplified for this model by taking a
percentage of the total amount of money generated from the ore. This amount is
initially in the range of 5% to 9.5% with an expected value of 7%.

The selling cost in each subsequent year is determined by applying a modifier to the previous year’s value. This modifier is a % change that is added to the previous year's cost using the formula $prev * (1 + mod)$. The selling cost modifier is in the range -0.1 to 0.45, with an expected value of 0.2275.

```{r selling_cost_config, results="hide", message=FALSE}
params$selling_cost_init <- list(a = 0.05, b = 0.095, c = 0.07)
params$selling_cost_delta <- list(a = -0.1, b = 0.45, c = 0.2275)
```

```{r selling_cost_profit}
env1$selling_cost <- gen_selling_rate(rounds = config$rounds, 
                                      years = config$years, 
                                      cost_init = params$selling_cost_init,
                                      cost_delta = params$selling_cost_delta)
```

```{r plot_selling_cost, echo=FALSE, fig.height=2}
env1$selling_cost %>%
  gather("Year", "Cost") %>%
ggplot(aes(x = Cost, fill = Year, colour = Year)) +
  geom_density(alpha = 0.6) +
  labs(title = "Selling Cost Distribution", 
       x = "Selling Cost (rate)",
       y = "Probability Density")
```

Selling costs are increasing over time, with increasing uncertainty.

### Equipment and Labour Costs

The mine requires fixed plant and employees to run. To simplify the model, all
these costs are rolled up into the range of \$2,000,000 to \$5,500,000 with an
expected value of \$4,000,000. As equipment can break down and even need
replacing this value is calculated each year and is independent of the previous
year.

```{r equip_labour_config, results="hide", message=FALSE}
params$equip_labour_range <- list(a = 2000000, b = 5500000, c = 4000000)
```

```{r equip_labour}
env1$equip_labour <- gen_independent(rounds = config$rounds, 
                                     years = config$years, 
                                     t = params$equip_labour_range)
```

```{r plot_equip_labour, echo=FALSE, fig.height=2}
env1$equip_labour %>%
  gather("Year", "Costs") %>%
ggplot(aes(x = Costs, fill = Year, colour = Year)) +
  geom_density(alpha = 0.6) +
  labs(title = "Equipment and Labour Cost Distribution", 
       x = "Costs (dollars)",
       y = "Probability Density")
```

The probability distribution for the equipment and labour costs for each year is
equal for all intents and purposes due to the fact that the assumptions are
independent from year to year.

### Exploration

The majority of the exploration parameters are related to decision making rather
than the environment, so they will be assessed in the next section of this 
report. As the second mine (if discovered) has a similar environment to the 
first mine, we can model it's environment here using the same approach as above.

```{r explore_env}
# Start by copying the things that are equal for both mines
env2 <- env1[c("gold_price", "interest_rate")]

# Gold production in this mine is independent, but with the same parameters
env2$gold_prod <- gen_gold_prod(rounds = config$rounds, 
                                years = config$years, 
                                prod_init = params$prod_init,
                                prod_delta = params$prod_delta)

# Selling cost in this mine is independent, but with the same parameters
env2$selling_cost <- gen_selling_rate(rounds = config$rounds, 
                                      years = config$years, 
                                      cost_init = params$selling_cost_init,
                                      cost_delta = params$selling_cost_delta)

# Equipment and labour costs are half that of the other mine, but not 
# independent. This means that they will depend on the simulation for mine 1.
env2$equip_labour <- env1$equip_labour / 2
```

For these values, the only one that is time-dependent is the gold production 
value, which effectively has a "time-shift" component based on the year the mine
is found. We can deal with this when we need to. All other values are considered
to be "fixed", i.e. the costs for year 3 do not depend on the year in which the
mine was discovered.

# Simulation and Decision Making

With the environment modelled appropriately, we can now commence simulation and
evaluation of possible decisions. 

### "Hands Off" Simulation

Firstly, let's look at some base-case assumptions for a first pass simulation:

* No attempt to find the second mine
* No mine sale activity

```{r}
profit_naive <- do_nothing_profits(env1)
```

Overall, how are the total profits in each year distributed?

```{r plot_naive, echo=FALSE, fig.height=2}
profit_naive$total_profit %>%
  gather("Year", "Profit") %>%
ggplot() +
  geom_density(aes(x = Profit, fill = Year, colour = Year), alpha = 0.6) +
  geom_line(data = data.frame(x=c(-4e6, -4e6), y=c(-Inf,Inf)), aes(x,y), col="red") +
  geom_line(data = data.frame(x=c(0, 0), y=c(-Inf,Inf)), aes(x,y), col="black") +
  labs(title = "Total Profit Distribution", 
       x = "Profit (dollars)",
       y = "Probability Density")
```

Without taking any action, the mining operation becomes more profitable over
time. Unfortunately there are `r sum(profit_naive$total_profit$Year1 < -4e6)`
cases where the profits are so far in the red that the bank will foreclose on
the mine - and it only gets worse over the five year period:

```{r echo=FALSE}
profit_naive$financial_position %>% gather(Year, Status) %>% table
```

Clearly there is some work to be done here!

### Controls Available

Each year, the operator of the mining company has the following options:

* Sell the leased mine (can only do this once)
* Explore for a third mine (can send 0-6 search parties, can only do this until found)
* Close the third mine (if it has been found)

This is significantly less control than a mining magnate would typically have
over their investments.

These decisions are to be made annually - no decisions can be made during the 
year. This means that there are 5 decision points - one decision point before 
each year.

### Visibility

Whilst the owner has models to inform the team about the expected value of
future environment variables (such as the price of gold, future production, etc)
they cannot see the future. This means that "informed" decisions within
individual simulations must be made according to the information available at
the current time in that simulation.

### Priorities

The primary objective is to avoid foreclosure from the bank. In the case that 
foreclosure cannot be avoided, the objective is to minimise the chance of 
foreclosure. For the purpose of this analysis, foreclosure is considered to be
*avoided* when less than 1% of simulations result in foreclosure.

The secondary objective is to maximise the median profit. Median has been 
selected because it is easy to reason about - there is a 50% chance that the 
profits will be at least this much.

The final objective is to minimise uncertainty, which will be measured using
*standard deviation* of the final profit.

Decisions will be modelled in isolation where possible (for example when
searching for a new mine) to reduce the complexity of the analysis and reduce 
the impact of componding assumptions in the model outputs.

### Naive approach to decisions

Looking simply at the decisions we can make prior to the start of the first
year, there are two decisions to make:

1. Should the second mine be sold?
1. Should the search for a third mine be funded? (and how much?)

Based on the "do nothing" simulation above we know that the chance of
foreclosure in the first year is negligible (0.1%), but the chance of 
foreclosure by the end of the 5 years is significant (39%). We should therefore
try to understand how our decisions in the first year impact this 5-year 
foreclosure probability.

#### Where are the profits coming from?

The profit/loss performance of the first mine is summarised below:

```{r echo=FALSE}
profit_naive$profit %>% 
  mutate(Year1 = case_when(Year1 < 0 ~ "Loss", Year1 >= 0 ~ "Profit"),
         Year2 = case_when(Year2 < 0 ~ "Loss", Year2 >= 0 ~ "Profit"),
         Year3 = case_when(Year3 < 0 ~ "Loss", Year3 >= 0 ~ "Profit"),
         Year4 = case_when(Year4 < 0 ~ "Loss", Year4 >= 0 ~ "Profit"),
         Year5 = case_when(Year5 < 0 ~ "Loss", Year5 >= 0 ~ "Profit")) %>% 
  gather(Year, Status) %>% 
  table
```

Overall, the first mine is problematic. We're really going to be leaning on the
second and third mines to help us turn a profit here.

The profit/loss performance of the second mine is summarised below:

```{r echo=FALSE}
profit_naive$leased_profit %>% 
  mutate(Year1 = case_when(Year1 < 0 ~ "Loss", Year1 >= 0 ~ "Profit"),
         Year2 = case_when(Year2 < 0 ~ "Loss", Year2 >= 0 ~ "Profit"),
         Year3 = case_when(Year3 < 0 ~ "Loss", Year3 >= 0 ~ "Profit"),
         Year4 = case_when(Year4 < 0 ~ "Loss", Year4 >= 0 ~ "Profit"),
         Year5 = case_when(Year5 < 0 ~ "Loss", Year5 >= 0 ~ "Profit")) %>% 
  gather(Year, Status) %>% 
  table
```

This is looking much healthier! We should still look at selling though, as the 
price in the first year is \$`r as.integer(params$sale_price$Year1)` which is 
far from negligible. We can look at the expected five-year profits and compare
these to the sale price to get a feel about whether this is a good deal.

```{r echo=FALSE}
profit_naive$leased_profit %>% 
  mutate(Total = Year1 + Year2 + Year3 + Year4 + Year5) %>% 
  ggplot() +
  geom_density(aes(x = Total), fill = "blue", alpha = 0.6) + 
  geom_line(data = data.frame(x=c(3e6, 3e6), y=c(-Inf,Inf)), aes(x,y), col="red") +
  labs(title = "Total Leased Profit Distribution", 
       x = "Profit (dollars)",
       y = "Probability Density")
```

The sale price is in no way a fair price based on the expected returns from the
second mine! However, the first objective is to avoid foreclosure, so we should
analyse this option to see if it has a positive impact on the future foreclosure
rate.

```{r}
sale_y1 <- profits_with_actions(env1, env2, params, sell_m2=1, find_m3=0)
```

```{r echo=FALSE}
sale_y1$financial_position %>% gather(Year, Status) %>% table
```

Whilst this decision completely avoids the risk of foreclosure in the first year
it makes the risk of foreclosure overall significantly worse due to the loss of
the income from the second mine in subsequent years.

> It appears to be a terrible idea to sell the second mine prior to the first
year of operation.

To assess whether it makes sense to search for the third mine at the start of
the first year, we need to understand the potential profits from that mine if we
were to find it.

```{r}
find_y1 <- profits_with_actions(env1, env2, params, sell_m2=0, find_m3=1)
```

```{r echo=FALSE}
find_y1$financial_position %>% gather(Year, Status) %>% table
```

This is a marginally better result than the sale of the second mine, but it's 
still no better than the "do nothing case".

> It does not appear to make sense to search for the second mine in year 1.

Finally, we should look at what happens when we sell the second mine and also
find the third mine in the first year.

```{r}
both_y1 <- profits_with_actions(env1, env2, params, sell_m2=1, find_m3=1)
```

```{r echo=FALSE}
both_y1$financial_position %>% gather(Year, Status) %>% table
```

Just as I thought, it's the worst of both worlds.

> You should not sell the second mine nor search for the third mine in the first
year.

### Robust approach to decisions

If we assume that the third mine can be found on demand and for zero cost (i.e. 
best case), then there are 30 different combinations of the following options:

* Sell the second mine never, or in one of the 5 years 
* Find the third mine never, or in one of the first 4 years 

For each of these options, we will compare the number of simulated foreclosures
in the final year, as minimising this number is the best way to achieve the 
first objective (avoid foreclosure).

```{r}
results <- tibble(
  sell_m2 = rep(NA, 30),
  find_m3 = rep(NA, 30),
  y5_foreclosure = rep(NA, 30)
)
for (i in 0:5) {
  for (j in 0:4) {
    idx <- i*5 + j + 1
    results[idx,"sell_m2"] <- i
    results[idx,"find_m3"] <- j
    profits <- profits_with_actions(env1, env2, params, sell_m2=i, find_m3=j)
    results[idx,"y5_foreclosure"] <- 
      sum(profits$financial_position$Year5 == "Foreclosure")
  }
}
rm(i, j, idx)
results %>% arrange(y5_foreclosure)
```

Clearly, in the absence of government intervention, there is no reason to take
any action at all, and purchasing the mine is a terrible idea in the first 
place.

# Making it Work

If Australian history tells us one thing, it's that mine operators know how to 
get things done. Given the challenging mining environment, it's not unreasonable
to expect that the following outcomes could be easily achieved with a little 
lobbying and a few cash-for-comments articles in The Australian:

1. In order to support Australian regional jobs, the tax will be set at 0%, 
which will halve the selling costs for the first and third mines.
1. Because Australian regional jobs cost a lot of money, the Government will 
relax the immigration laws to allow foreign labour imports. This will halve the
total equipment and labour costs for the first mine, and this will also flow in 
to reduced costs for the third mine

We can model these new assumptions, and see how they impact the analysis:

```{r}
env3 <- env1 # Make a copy
env4 <- env2 # Make a copy

env3$selling_cost <- env1$selling_cost / 2
env3$equip_labour <- env1$equip_labour / 2
env4$equip_labour <- env3$equip_labour / 2

profits_with_actions(env3, env4, params, sell_m2=0, find_m3=0)$financial_position %>% 
  gather(Year, Status) %>% 
  table()
```

Now we're talking! Let's take a look at what actions will help drive the chance
of foreclosure to zero:

```{r}
results <- tibble(
  sell_m2 = rep(NA, 30),
  find_m3 = rep(NA, 30),
  y5_foreclosure = rep(NA, 30)
)
for (i in 0:5) {
  for (j in 0:4) {
    idx <- i*5 + j + 1
    results[idx,"sell_m2"] <- i
    results[idx,"find_m3"] <- j
    profits <- profits_with_actions(env3, env4, params, sell_m2=i, find_m3=j)
    results[idx,"y5_foreclosure"] <- 
      sum(profits$financial_position$Year5 == "Foreclosure")
  }
}
rm(i, j, idx)
results %>% arrange(y5_foreclosure)
```

There is a bit of variability here, ranging from less 0.01% foreclosure to
almost 0.1% foreclosure. It's definitely unlikely! There is now also a strong
signal that finding the second mine as soon as possible is a good thing to do.
We should start to incorporate the cost of the search - starting with 6 search
parties to check everything is working.

```{r}
params$search_price <- list(   0, 2e5L, 5e5L, 9e5L, 13e5L, 17e5L) 
params$search_prob  <- list(0.09, 0.28, 0.39, 0.48,  0.56,  0.62) 

profits_with_search(env3, env4, params, sell_m2=0, search=6)$financial_position %>% 
  gather(Year, Status) %>% 
  table()
```

It looks like we have some balancing to do! Let's look at all of the options 
together.

```{r}
results <- tibble(
  sell_m2         = rep(NA, 36),
  search_strength = rep(NA, 36),
  y5_foreclosure  = rep(NA, 36)
)
for (i in 0:5) {
  for (j in 1:6) {
    idx <- i*6 + j
    results[idx,"sell_m2"] <- i
    results[idx,"search_strength"] <- j
    profits <- profits_with_search(env3, env4, params, sell_m2=i, search=j)
    results[idx,"y5_foreclosure"] <- 
      sum(profits$financial_position$Year5 == "Foreclosure")
  }
}
rm(i, j, idx)
results %>% arrange(y5_foreclosure)
```

From this analysis it is very clear that searching is worthwhile, but that a 
single search party is best when considering the risk of foreclosure. 

Finally, let's consider how profits are impacted with each of these decisions.

```{r}
results <- tibble(
  sell_m2         = rep(NA, 36),
  search_strength = rep(NA, 36),
  y5_foreclosure  = rep(NA, 36),
  median_cash_y5  = rep(NA, 36)
)
for (i in 0:5) {
  for (j in 1:6) {
    idx <- i*6 + j
    results[idx,"sell_m2"] <- i
    results[idx,"search_strength"] <- j
    profits <- profits_with_search(env3, env4, params, sell_m2=i, search=j)
    results[idx,"y5_foreclosure"] <- 
      sum(profits$financial_position$Year5 == "Foreclosure")
    results[idx, "median_cash_y5"] <- median(profits$cash$Year5[[1]])
  }
}
rm(i, j, idx)
results %>% arrange(desc(median_cash_y5))
```

This is great news - there is a really clear strategy emerging here:

1. Pressure the government for tax breaks, and hire cheap foreign workers. This is essential, and the mine should not be purchased without these conditions.
1. Do not sell mine 2, and if you do have to sell it (for example if you have cashflow issues), sell it as late as possible.
1. Send a small search party of only 1 or 2 groups.

**If you follow these three steps, then there is less than 0.01% chance of the 
bank foreclosing on the mine, and there is a 50% chance of making more than $12
million in cash by the end of year 5.**

# Reproducability

This analysis was performed using R, and the environment is described below to 
allow the analysis to be reproduced faithfully.

### Session info

```{r colophon, echo=FALSE}
devtools::session_info()
```
